---
title: "Maps"
output: html_document
date: "2024-01-04"
author: Masha Faticov
---

Upload packages
```{r}
# Upload libraries - check if you need to install some of the packages
library(tidyverse) 
library(magrittr) # pipes
library(lintr) # code linting
library(sf) # spatial data handling
library(cowplot) # stack ggplots
library(readxl) #package to read excel files in R
library(ggmap) #package to download maps from google maps
library(ggsn) #package to deal with spatial data
library(rgdal)
library(proj4)  
library(unikn)  
library(ggspatial)

```
Steps to create maps:

1. We need to find and load background map to R
 -  region boundaries can be stored in shapefiles or geoJSON files (use package SF) 
 - some R libraries also provide the data for the most common places, especially boundries of countries (map, mapdata packages - a few examples)
 -  is also possible to use google maps to load and style backgrounds (found it to be the easiest option for me) - package ggmap and I will show this example
 
 2. We also need data to plot on the map
 - easiest if you have coordinates (Google coordinates WGS 84 will work) for your data to plot on the map
 - we will use ggplot2 and ggmap package to plot data
 
 Let's plot a map
 
```{r}
#Load data
dataformaps <- read_excel("C:/Users/maria/Dropbox/Postdoc_Sherbrooke/New Frontiers/Maps_NF_all_traps/Montreal.xlsx")

#check whether variables are numeric or character - transform the character ones to numeric
dataformaps$Lat <- as.numeric(as.character(dataformaps$Lat))
dataformaps$Long <- as.numeric(as.character(dataformaps$Long))

#Access Googple maps
#to access maps in Googple maps you need API key. Call it using the key below:
register_google(key = "AIzaSyB7ujAyTHJ5NCnRKyfz6YRjDzr9JEEboyo")

# get map from Google maps by providing coordinates of the center of the city
map1 = get_googlemap(center = c(-73.616986, 45.489037), color='bw', maptype='terrain', 
 zoom=10,style = list(c(element="labels", visibility="off"),
 c("&style=", feature="road.arterial", color="0xffffff", visibility="off"),
 c("&style=", feature="road.highway", color="0xffffff", visibility="off"))) 

#center = c(-73.616986, 45.489037): Specifies the longitude and latitude around which to center the map.
#color='bw': Indicates that the map should be in black and white
#zoom=10: Sets the zoom level of the map. The higher the number, the closer the zoom
#style: A list of styles to further customize the appearance of the map. In this case:
#c(element="labels", visibility="off"): Turns off all labels on the map (like street names and points of interest).
#c("&style=", feature="road.arterial", color="0xffffff", visibility="off"): Tries to turn off arterial roads, setting their color to white (0xffffff), though the syntax appears to be incorrect.
#c("&style=", feature="road.highway", color="0xffffff", visibility="off"): Tries to turn off highways, also setting their color to white, with a similar issue in the syntax.

p_map1 <- ggmap(map1, extent="panel") +
  scale_x_continuous(limits = c(-74.0, -73.42), expand = c(0, 0)) +
  scale_y_continuous(limits = c(45.36, 45.70), expand = c(0, 0)) +
  geom_point(data=dataformaps, aes(x=Long, y=Lat, size=2,color="blue", fill="blue")) +
  annotate("text", x=-73.6, y=45.432, label= "Montréal") +
  theme(
    legend.position = "right",
    text = element_text(color = "#22211d"),
    plot.background = element_rect(fill = "white", color = NA), 
    panel.background = element_rect(fill = "white", color = NA),
    axis.title.y = element_text(vjust = +3),
    axis.title.x = element_text(vjust = -0.30),
    legend.background = element_rect(fill = NA, color = NA),
    legend.key=element_rect(fill= NA),
    legend.title = element_text(face = "bold"),
    axis.line = element_line(colour = "black"), 
    panel.border = element_rect(colour = "black", fill=NA, linewidth=1), 
    plot.title = element_text(size= 16, hjust=1, color = "#4e4d47", margin = margin(b =  
    0.1, t = 0.4, l = 2, unit = "cm"))) + 
    guides(color = guide_legend(override.aes = list(size = 5))) +
    labs(x = "Latitude",y = "Longitude", caption = "Population size: 1 760 000 \n Population density per square kilometre: 4830") 

p_map1 + 
annotation_north_arrow(location = "tl", which_north = "true", pad_x = unit(0.0, "in"), pad_y = unit(0.2, "in"),style = north_arrow_minimal) + 
ggsn::scalebar(x.min=-74.0,x.max=-73.48,y.min=45.38,y.max=45.73, transform=TRUE, location = "bottomright",dist_unit = "km", dist=6, st.bottom = TRUE, st.size = 2)
p_map1

```
 
 
 
