---
title: "Introduction to bar/column plots with ggplot2."
date: "2024-01-06"
Author: Jonathan Rondeau-Leclaire
---

```{r, include=FALSE}
library(tidyverse)
```

## Data preparation

The data : Microbial family abundance by sample.

We have 131 samples taken on four different plant hosts (mosses) on either the brown or green compartment of the plant. For each microbial Family identified, the count represents the absolute number of sequences in a sample that were attributed to that Family. We want to compare these numbers across compartments and hosts.

First, let's have a look at the data: 

```{r}
# Import the data file
seq_count <- read.delim("../../Data/sequence_counts.tsv") # by default, read.delim assumes the file has a header, and that it is separated by tabulations ("\t")
head(seq_count)
# How many samples do we have by group and compartment ?
seq_count %>% 
  select(Sample, Host, Compartment) %>% # ignore the seq_count and Family columns
  unique %>% # Keep one line per unique combinations of these variables
  dplyr::summarise(n_samples = n(), # count the number of samples for each
                   .by = c('Host', 'Compartment'))
```
The columns whose values are not numbers are considered "character" columns. You can see this by looking at the structure of your object, where "chr" means character and "int" means integer.

```{r}
str(seq_count)
```
To work well with ggplot, we need them to be factors. You can see factors as a set of categories that represents each unique value in a variable. For example, we have two Compartment factors (Brown or Green). Here's a quick way to convert all character columns to factors, using a little trick from the very useful 'dplyr' package.

```{r}
seq_count <- seq_count %>% 
  # the mutate() function modifies the values of a variable
  mutate(across(where(is.character), as.factor))

# Make sure the factors are as expected
str(seq_count)
```

There, you can see that our four variables are now factors and you can validate the number of levels for each. 

## Barplot : Family 

We want to look at how microbial families are distributed across our samples using their sequence counts. We'll start very simple, by using the mean of all samples for each Family. For this, we need to do a short data manipulation, because right now our dataset has one observation per family _per sample_. 

```{r}
# Dataframe
family.df <- seq_count %>% 
  group_by(Family) %>% 
  summarise(mean_count = mean(seq_count))

# Plot 
ggplot(family.df, 
       aes(x = mean_count, y = Family)
       ) + 
  geom_col() 
```

That's nice, but now we'd like to see if these counts differ much across compartments, i.e. the green vs. the brown part of the plants. Therefore, we need to calculate the mean for each family. We'll also compute the standard deviation of those means for later.

```{r}
# Dataframe
familyComp.df <- seq_count %>% 
  group_by(Family, Compartment) %>% 
  summarise(mean_count = mean(seq_count),     
            sd_count = sd(seq_count),        # calculate SD
            .groups = 'drop')                # drops the grouping structure

# Plot
ggplot(familyComp.df, aes(x = mean_count, y = Family, 
                          fill = Compartment)) + # fill by compartment 
  geom_col() 
```

That's not bad, but we might want to have each column side-by-side instead of stacked on top of each other. The "position" argument is going to help us here. While we're at it, let's use some proper colours:

```{r}
ggplot(familyComp.df, aes(x = mean_count, y = Family, fill = Compartment)) + 
  geom_col(position = "dodge") # Dodge the columns, so we can read the counts
```

That's better. However, it could be useful to show how precise these means are. We can add the standard deviation of each as an error bar. Look carefully, we need to modify the "position" argument a bit so everything aligns properly.

```{r}
ggplot(familyComp.df, aes(x = mean_count, y = Family, fill = Compartment)) + 
  geom_col(position = position_dodge(width = 1)) + # slight modification
  geom_errorbar(aes(xmin = mean_count - sd_count, 
                    xmax = mean_count + sd_count),
                width = 0.3, # size of the "caps"
                linewidth = 0.2, # thickness
                position = position_dodge(width = 1)) # same as in geom_col
```

Since plots are meant to be pretty, let's do a couple more things: 

1. Use some coherent colours (green and brown!)
2. Remove the gray background (personal preference) by imposing a minimal theme
3. Rename the x axis and add a plot title
4. Remove the horizontal grid
5. Space-out the column pairs vertically to better distinguish families 
6. Use a different font size and police

Since this will be our final plot for this exercise, we'll also save it into an object in order to be able to export it. 

```{r}
comp_colours <- c("lightsalmon4", "springgreen4")

myplot <- # save this plot as an object
  ggplot(familyComp.df, aes(x = mean_count, y = Family, fill = Compartment)) + 
  geom_col(position = position_dodge(width = 0.7), # cluster columns by family
           width = 0.7) + # readjust the columns so they don't overlap
  geom_errorbar(aes(xmin = mean_count - sd_count, xmax = mean_count + sd_count),
                width = 0.3, linewidth = 0.2, position = position_dodge(width = 0.7)) +
  scale_fill_manual(values = comp_colours) + # follows factor levels
  labs(title = "Family sequence counts by compartment.", # Title
       x = "Mean of sample sequence counts") + # meaningful label
  theme_minimal(base_family = "Baskerville", # new font, and this theme removes gray background
                base_size = 12) + # font size
  theme(panel.grid.major.y = element_blank()) # remove horizontal grid
myplot
```

Saving this plot can be easily achieved through ggsave, which provides a straightforward way 
```{r}
ggsave(filename = "./barplot.jpg", plot = myplot, 
       width = 1600, height = 900, units = 'px', dpi = 150)
```

## Exercise

Try to create a barplot with mean seq_count _by Host_, without specifying colours. Try applying a different theme (hint: type ?theme_ in your console and a semi-automatic selection appears with many themes to play around with). Add a caption to your plot to specify that we are using 131 samples, and change the legend label to something more explicit than "host". (hint: look at the labs() functions). Do not add error bars

```{r, include = FALSE}
seq_count %>% 
  group_by(Family, Host) %>% 
  summarise(mean_count = mean(seq_count),     
            sd_count = sd(seq_count),        # calculate SD
            .groups = 'drop') %>% 

ggplot(aes(x = mean_count, y = Family, fill = Host)) + 
  geom_col(position = position_dodge(width = 0.8), # cluster columns by family
           width = 0.8) + # readjust the columns so they don't overlap
  labs(title = "Family sequence counts by compartment.", # Title
       x = "Mean of sample sequence counts",
       caption = "n = 131 samples from the Québec Boreal Forest.",
       fill = "Host moss species") + # meaningful label
  theme_light(base_family = "Baskerville") +
  theme(plot.caption.position = "plot", # sends it all the way to the right.
        plot.title.position = "plot") # Sends it all the way to the left
```

## Barplots with facets

A different way of looking at the data could be to compare Host and Compartment at the same time. To visualize the families across these two sample data, we can use faceting. Notice that we are back to the original (non-summarised) data.

In this case, we'll start by plotting the counts for each compartment

```{r}
ggplot(seq_count, aes(y = Compartment, x = seq_count)) + 
  geom_col(aes(fill = Family))
      # Notice: we don't use the "dodge" option, so the counts stack.
```

How can we look at each host separately? Faceting will help us !

```{r}
ggplot(seq_count, aes(y = Compartment, x = seq_count)) + 
  geom_col(aes(fill = Family)) +
    # This is where the magic happens:
  facet_grid(rows = vars(Host), # Adds facets
             scales="free") # allows different x-axis scales for each facet
```

That's fine, but it reveals a weakness in the way we are presenting data. Indeed, the number of sequences is usually much smaller in the green compartment. One way to solve this is to normalize to 1, which is equivalent to showing the relative counts instead of the absolute counts. We can easily achieve this without modifying our original data:

```{r}
ggplot(seq_count, aes(y = Compartment, x = seq_count)) + 
  geom_col(aes(fill = Family), 
           position = 'fill') + # normalises total values to 1
  facet_grid(rows = vars(Host)) +
  xlab("Proportion of sequence counts") # not absolute values
```

Notice, the counts were normalized to 1.