---
title: "barplots_ggplot"
date: "2024-01-06"
Author: Jonathan Rondeau-Leclaire
---

```{r, include=FALSE}
library(tidyverse)
```

The data : Microbial families abundance by sample.
We have 131 samples taken on four different plant hosts (mosses) on either the brown or green compartment of the plant. For each microbial family identified, the count represents the absolute number of sequences in a sample that were attributed to that family. We want to compare these numbers across compartments and hosts.

First, let's have a look at the data: 

```{r setup}
(seq_count <- read.delim("sequence_counts.tsv"))
```

As you can see below the column names, the ones whose values are not numbers are considered "character" columns. To work well with ggplot, we need them to be factors. You can see factors as a set of categories that represents each unique value in a variable. For example, we have two Compartment factors (Brown or Green). Here's a quick way to convert all character columns to factors, using a little trick from the very useful 'dplyr' package.










