---
title: "" #ECL750<br>François Rousseu<br>Hiver 2021
author: ""
date: ""
output:
  html_document:
    depth: 5
    fig_height: 5
    fig_width: 6
    number_sections: no
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
      smooth_scroll: no
editor_options: 
  chunk_output_type: console
---

```{=html}
<script>
    $(document).ready(function() {
      $items = $('div#TOC li');
      $items.each(function(idx) {
        num_ul = $(this).parentsUntil('#TOC').length;
        $(this).css({'text-indent': num_ul * 20, 'padding-left': 0});
      });

    });
</script>
```


```{=html}
<style>

pre.r {
    background-color: #EEEEEE;
    border-color: #EEEEEE;
    font-size: 12pt;
}

pre code {
  font-size: 12pt;
}

body {
  font-size: 14pt;
  margin-right: 5vw;
  margin-left: 0vw;
  /* max-width: 80vw; */
}

.main-container {
    max-width: 100vw !important;
} 

#TOC {
  font-size: 0.85vw;
  border-color: white;
  max-width: 15vw;
  margin-top: 5vh;
}

.list-group-item.active:focus{
    z-index: 2;
    color: darkgreen;
    background-color: #EEEEEE;
    /* font-weight: bolder; */
    text-decoration: underline;
}

.list-group-item.active:hover {
    z-index: 2;
    color: darkgreen;
    background-color: #FFFFFF;
    /* font-weight: bolder; */
    text-decoration: underline;
}

.list-group-item.active{
    z-index: 2;
    color: darkgreen;
    background-color: #FFFFFF;
    /* font-weight: bolder; */
    text-decoration: underline;
}

h1.title {
  margin-top: 50px;
  font-size: 42px;
  color: DarkGreen;
  font-weight: bold;
}

h1, h2, h3, h4, h5, h6 {
  margin-top: 5vh;
  padding-top: 40px;
  padding-bottom: 20px;
  color: DarkGreen;
  font-weight: bold;
}

h1, h2, h3, h4, h5 {
  font-size: 42px;
}


.pres { /* write {.pres} next to each section to apply css  */
   padding-top: 100px;
   margin-bottom: 700px;
}

/* html{
  scroll-behavior: smooth;
} */


</style>
```


```{r setup, include=FALSE}
library(scales)
knitr::opts_chunk$set(echo=TRUE, tidy=TRUE, error=TRUE, fig.align="center", fig.width=8, fig.height=6, cache=FALSE, message=FALSE)
options(width=150)
```

# Visualizing models with ggeffects

The package [ggeffects](https://strengejacke.github.io/ggeffects/index.html) is for visualizing results of analyses and model-based predictions. It is based on [ggplot2](https://ggplot2.tidyverse.org/index.html)

<!-- <iframe src="https://strengejacke.github.io/ggeffects/index.html" height="400" width="1000" title="Iframe Example"></iframe> -->

```{r, message = FALSE, include = FALSE, echo = FALSE}
d <- read.csv("https://raw.githubusercontent.com/frousseu/ECL749/master/data/germination.csv", header = TRUE)
write.csv(d,"/home/frousseu/Documents/github/graphs_uds_2024/data/germination.csv",row.names=FALSE)
```

## Plot raw data

```{r, out.width="60%", fig.height=4,fig.width=6, fig.align="center", message=FALSE, echo=TRUE}
library(ggplot2)

head(iris)

ggplot(iris,aes(Species,Petal.Length))+
  geom_jitter(width = 0.1, alpha = 0.2) + 
  theme_light()

```

## An ANOVA / linear model

Raw data is good, but a model is (usually) better!

```{r, out.width="60%", fig.height=4,fig.width=6, fig.align="center", message=FALSE, echo=TRUE}

m <- lm(Petal.Length~Species,data=iris)
anova(m)

```

## Enters ggeffects!

```{r, out.width="60%", fig.height=4,fig.width=6, fig.align="center", message=FALSE, echo=TRUE}

library(ggeffects)

p <- ggpredict(m,terms=c("Species"))

plot(p)

```

## Add data

Model predictions are good, but it is even better when they are plotted along with the data!

```{r, out.width="60%", fig.height=4,fig.width=6, fig.align="center", message=FALSE, echo=TRUE}

plot(p, add.data = TRUE, jitter = TRUE, dot_alpha = 0.15, color="darkgreen") 

```

## Add more informative labels

```{r, out.width="60%", fig.height=4,fig.width=6, fig.align="center", message=FALSE, echo=TRUE}

plot(p, add.data = TRUE, jitter = TRUE, dot_alpha = 0.35, color="darkgreen") +
  labs(
    x = "Iris species",
    y = "Petal lengths (cm)",
    title = "Petal lengths of iris species"
  )
```


## What does ggpredict produces ?

```{r, out.width="60%", fig.height=4,fig.width=6, fig.align="center", message=FALSE, echo=TRUE}

ggpredict(m,terms=c("Species"))

data.frame(ggpredict(m,terms=c("Species")))

```


## A data frame with the values to (gg)plot

```{r, out.width="50%", fig.height=3,fig.width=4, fig.align="center", message=FALSE, echo=TRUE}

x <- data.frame(ggpredict(m,terms=c("Species")))
x

ggplot(x, aes(x, predicted)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high),size = 0.2) +
  geom_jitter(data = iris, aes(Species, Petal.Length), alpha = 0.1) +
  theme_bw()

```


## Hummingbirds

```{r, fig.height = 4, fig.width = 11, fig.align="center" }
d <- read.csv("https://raw.githubusercontent.com/isabelle-laforest/R_UdeS_2024/main/Data/colibris.csv")
d <- na.omit(d)

cols <- c("sexe","age")
d[cols] <- lapply(d[cols],factor)

# A glm and poisson family for count data (number of parasites)
m <- glm (parasites ~ sexe + age + masse, data = d, family = poisson) 
```


## Plot predictions
```{r, fig.height = 4, fig.width = 11, fig.align="center" }

library(magrittr) # for the %>% 
library(patchwork)

g1 <- ggpredict(m, terms=c("masse")) %>% 
  plot(add.data = TRUE, jitter = 0.3, dot_alpha = 0.1)

g2 <- ggpredict(m, terms=c("sexe")) %>% 
  plot(add.data = TRUE, jitter = 0.3, dot_alpha = 0.1)

g3 <- ggpredict(m, terms=c("age")) %>% 
  plot(add.data = TRUE, jitter = 0.3, dot_alpha = 0.1)

g1 + g2 + g3

```


## Predictions combined
```{r, fig.height = 4, fig.width = 7, fig.align="center"}

g <- ggpredict(m, terms=c("masse", "sexe", "age"))
plot(g, add.data = TRUE, jitter = 0.3, dot_alpha = 0.5)

```

<!--

## Tree plantation - Anova
```{r, out.width="80%", fig.height=4, fig.width=8, fig.align="center", message=FALSE,}
d <- read.csv("https://raw.githubusercontent.com/isabelle-laforest/R_UdeS_2024/main/Data/pines.csv", header = TRUE)
head(d,3)

cols <- c("deer", "fert", "cover", "spacing")
d[cols] <- lapply(d[cols], factor)

m <- lm (height~fert*cover+height_ini,data=d)
anova(m)
```


## Tree plantation - Graph
```{r, out.width="80%", fig.height=4, fig.width=8, fig.align="center", message=FALSE,}

g1 <- ggpredict(m, terms=c("height_ini")) %>% 
  plot(add.data=TRUE, jitter = 0.1, dot_alpha = 0.1)

g2 <- ggpredict(m, terms=c("fert", "cover")) %>% 
  plot(add.data=TRUE, jitter = 0.1, dot_alpha = 0.1)

layout <- "
AAABB
"  
g1 + g2 + plot_layout(design = layout)

```

-->

## More complex examples

```{r, out.width="80%", fig.height=4,fig.width=8, fig.align="center", message=FALSE, echo=TRUE}

d<-read.csv("https://raw.githubusercontent.com/isabelle-laforest/R_UdeS_2024/main/Data/germination.csv")
head(d)

# A glm with a binomial family for binomial data (logistic regression, germination or not 0/1)
m<-glm(germ~esp*substrat*frequence,data=d,family=binomial)
```


## Plot the % of germination
```{r, out.width="80%", fig.height=4,fig.width=8, fig.align="center", message=FALSE, echo=TRUE}
# predictions from model
g<-ggpredict(m,terms=c("esp","substrat","frequence"))

plot(g)

```



## Add observations ?

<!--

```{r, out.width="80%", fig.height=4,fig.width=8, fig.align="center", message=FALSE, echo=TRUE}

plot(g, add.data=TRUE, jitter=0.05, dot_alpha=0.2)


```

-->

## Challenge

Plot a graph from a model that looks at the effect of height (_hauteur_), exposition (_exposition_) and beach type (_Type_) on the number of corals species (_richesse_).

```{r, out.width="80%", fig.height=4,fig.width=8, fig.align="center", message=FALSE, echo=TRUE}

d <- read.csv("https://raw.githubusercontent.com/frousseu/ECL749/master/data/plages.csv")
h <- read.csv("https://raw.githubusercontent.com/frousseu/ECL749/master/data/plage_type.csv")
d <- merge(d, h, by.x = "plage", by.y = "Plage")
head(d)

m <- glm(richesse ~ hauteur + exposition + Type, data = d, family = poisson)

```


<!--

### Solution

```{r, out.width="80%", fig.height=4,fig.width=8, fig.align="center", message=FALSE, echo=TRUE}




# predictions from model
g<-ggpredict(m,terms=c("hauteur","exposition","Type"))

plot(g, dot_size=1, add.data = TRUE)

```

-->

<!-- ## Owls
```{r, out.width="80%", fig.height=4,fig.width=8, fig.align="center", message=FALSE, echo=TRUE}
library(ggeffects)
library(ggplot2)

d<-read.csv("https://raw.githubusercontent.com/frousseu/ECL749/master/data/chouette.csv")
head(d)

d["year"]<-lapply(d["year"],factor)

m<-glm(CRpres~for_feuil*year,data=d,family=binomial)

# predictions from model
g1<-ggpredict(m,terms=c("for_feuil")) |> plot()
    
g2<-ggpredict(m,terms=c("year")) |> plot()

g2<-ggpredict(m,terms=c("for_feuil","year")) |> plot()

g1 + g2

```

-->

<!--

## Exercice

```{r}
d <- read.csv("https://raw.githubusercontent.com/isabelle-laforest/R_UdeS_2024/main/Data/pines.csv", header = TRUE)
head(d,3)

cols <- c("deer", "fert", "cover", "spacing")
d[cols] <- lapply(d[cols], factor)

m <- lm (height~fert*cover,data=d)
anova(m)

g <- ggpredict(m, terms=c("fert", "cover"))
plot(g, add.data=TRUE, jitter = 0.1, dot_alpha = 0.15)

#hypothesis_test(m, c("fert", "spacing"))

#ggplot(g, aes(x, predicted)) +
#  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), position = position_dodge(width=0.05)) +
#  geom_point(data=d,aes(fert,height), position = position_dodge(width=0.05))
  

```

-->






