---
title: "box_plot_ggplot2"
date: "2024-01-03"
Author: Masha Faticov
---

Load data:
- moss bacteria from Norther Quebec, 
- measured bacterial richness from 4 different moss species and two compartments
- estimated diversity and richness of bacteria

```{r setup, include=FALSE}
moss_data <- read.csv("C:/Users/maria/Dropbox/Postdoc_Sherbrooke/Courses/R course visualization with ggplot/Data/mossdata.csv", sep=';', header =TRUE)
```

Question of the study:

Does bacterial diversity differ between moss species and moss compartments?
One way to visualize the answer to this question is to use box plots


Let's start with a simple box plot
```{r}
plot1 <- ggplot(moss_data, aes(x = factor(Species), y = Shannon)) +
  geom_boxplot() +
  xlab("Moss species") +
  ylab("Bacterial diversity (Shannon index)") +
  ggtitle("Box plot of bacterial diversity among moss species")
plot1
```
What do we see on this plot?
- Box itself shows the interquantile range - the middle (50%) of the data or median
- Data points that fall outside 1.5 times the IQR from the quartiles are potentially outliers

 Data distribution is hidden behind each box. For instance, a normal distribution could look exactly the same as a bimodal distribution. Let's plot together with data points:
 
```{r}
plot2 <- ggplot(moss_data, aes(x = factor(Species), y = Shannon)) +
  geom_boxplot() +
  geom_jitter(color="black", size=0.4, alpha=0.9) +
  xlab("Moss species") +
  ylab("Bacterial diversity (Shannon index)") +
  ggtitle("Box plot of bacterial diversity among moss species")
plot2

#change shape of the boxplot - notch and notchwidth - make it easier to compare medians
plot2a <- ggplot(moss_data, aes(x = factor(Species), y = Shannon)) +
  geom_boxplot(notch=TRUE,
        notchwidth = 0.9,
        # custom outliers
        outlier.colour="red",
        outlier.fill="red",
        outlier.size=3) +
  geom_jitter(color="black", size=0.4, alpha=0.9) +
  xlab("Moss species") +
  ylab("Bacterial diversity (Shannon index)") +
  ggtitle("Box plot of bacterial diversity among moss species")
plot2a
```
 
 
We have three different species and we would like to show it on our box plot? 
There are several ways to do it:

- first you can pick and assign color yourself
#https://derekogle.com/NCGraphing/resources/colors
#https://r-graph-gallery.com/42-colors-names.html

- using color brewers in R (packages that pre made color schemes for you), e.g., RColorBrewer: sequential palettes, diverging and qualitative
#https://r-graph-gallery.com/38-rcolorbrewers-palettes.html 
#https://github.com/EmilHvitfeldt/r-color-palettes

install.packages("RColorBrewer")
library(RColorBrewer)
 
```{r}
#1. Picking color ourselves - fill vs color (contour)
#alpha --  geom_boxplot(alpha = 0.8)
plot3a <- ggplot(moss_data, aes(x = factor(Species), y = Shannon, fill = factor(Species))) +
  geom_boxplot(alpha = 0.8) +
  geom_jitter(color="black", size=0.4, alpha=0.9) +
  scale_fill_manual(values = c("lightpink4", "slategray4","plum3", "darkseagreen4")) +
  xlab("Moss species") +
  ylab("Bacterial diversity (Shannon index)") +
  ggtitle("Box plot of bacterial diversity among moss species")
plot3a

#2. Picking color ourselves 
plot3b <- ggplot(moss_data, aes(x = factor(Species), y = Shannon, fill = factor(Species))) +
  geom_boxplot() +
  geom_jitter(color="gray30", size=0.5, alpha=0.6) +
  xlab("Moss species") +
  ylab("Bacterial diversity (Shannon index)") +
  ggtitle("Box plot of bacterial diversity among moss species") +
  scale_fill_brewer() 
plot3b

#color instead of fill
plot3c <- ggplot(moss_data, aes(x = factor(Species), y = Shannon, color = factor(Species))) +
  geom_boxplot() +
  geom_jitter(color="gray30", size=0.5, alpha=0.6) +
  xlab("Moss species") +
  ylab("Bacterial diversity (Shannon index)") +
  ggtitle("Box plot of bacterial diversity among moss species") +
  scale_color_brewer() 
plot3c

plot3d <- ggplot(moss_data, aes(x = factor(Species), y = Shannon, fill = factor(Species))) +
  geom_boxplot() +
  geom_jitter(color="gray30", size=0.5, alpha=0.6) +
  xlab("Moss species") +
  ylab("Bacterial diversity (Shannon index)") +
  ggtitle("Box plot of bacterial diversity among moss species") +
  scale_fill_brewer(palette = "Set1") 
plot3d
```

What if we want to also demonstrate mean value (avergaes) and not only median? We can do that
```{r}
#add mean to the graph -  stat_summary(fun.y=mean, geom="point", shape=20, size=14, color="", fill="") 
plot4 <- ggplot(moss_data, aes(x = factor(Species), y = Shannon, fill = factor(Species))) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = c("lightpink4", "slategray4","plum3", "darkseagreen4")) +
  stat_summary(fun=mean, geom="point", shape=20, size=3, color="royalblue4") +
  geom_jitter(color="black", size=0.4, alpha=0.9) +
  xlab("Moss species") +
  ylab("Bacterial diversity (Shannon index)") +
  ggtitle("Bacterial diversity of different moss species") 
plot4
```

Let's look at our data again:
View(moss_data)

We have moss species, but we also are interested how bacterial diversity differ between moss compartments and whether this pattern is similar or different among moss species

Let's plot this!

Useful links: #https://ggplot2.tidyverse.org/reference/theme.html

```{r}
View(moss_data)

#grouped
plot5 <- ggplot(moss_data, aes(x = factor(Species), y = Shannon, fill = factor(Compartment))) +
  geom_boxplot(alpha = 0.9) +
  scale_fill_manual(values = c("salmon4", "darkolivegreen"), labels=c("Brown","Green")) +
  #Adds individual data points on the box plots with a jittered position to avoid overplotting.
#position_jitterdodge ensures that the points are spread out within each category for clarity.
#alpha=0.2 sets the transparency of the points.
  geom_point(position=position_jitterdodge(dodge.width=0.7, seed=0.5),alpha=0.4,show.legend = FALSE) +
  xlab("Moss species") +
  ylab("Bacterial diversity (Shannon index)") 
plot5

#faceting - one facet per species 
plot5b <- ggplot(moss_data, aes(x = factor(Species), y = Shannon, fill = factor(Compartment))) +
  geom_boxplot(alpha = 0.9) +
    facet_wrap(~ Species, scale="free") +
  scale_fill_manual(values = c("salmon4", "darkolivegreen"), labels=c("Brown","Green")) +
  geom_point(position=position_jitterdodge(dodge.width=0.7, seed=0.5),alpha=0.4,show.legend = FALSE) +
  #xlab("Moss species") +
  ylab("Bacterial diversity (Shannon index)") 
plot5b

#we need to modify the THEME of our plot 
#THEME is a set of aesthetic choices that affect the non-data display of a plot. This includes the display of background, grids, axis, ticks, labels, title, legend, and other non-data elements
#All these changes can be done in the THEME
plot5c <- ggplot(moss_data, aes(x = factor(Species), y = Shannon, fill = factor(Compartment))) +
  geom_boxplot(alpha = 0.9) +
  scale_fill_manual(values = c("salmon4", "darkolivegreen"), labels=c("Brown","Green")) +
  geom_point(position=position_jitterdodge(dodge.width=0.7,    
  seed=0.5),alpha=0.4,show.legend = FALSE) +
  ylab("Bacterial diversity (Shannon index)")+
  facet_wrap(~ Species, scale="free") +
  theme(axis.text.x = element_blank())
plot5c

#In the THEME we can:
#1. change ext Size and Font: 
plot5d <- ggplot(moss_data, aes(x = factor(Species), y = Shannon, fill = factor(Compartment))) +
  geom_boxplot(alpha = 0.9) +
  scale_fill_manual(values = c("salmon4", "darkolivegreen"), labels=c("Brown","Green")) +
 geom_point(position=position_jitterdodge(dodge.width=0.7, seed=0.5),alpha=0.4,show.legend = FALSE) +
  ylab("Bacterial diversity (Shannon index)")+
  facet_wrap(~ Species, scale="free") +
  theme(axis.text.x = element_blank())+
  theme(text = element_text(size = 12),
  axis.title = element_text(size = 13, face = "bold"),
  axis.text = element_text(size = 10))
 plot5d

 #2. MOdify legend position and titles
 
 plot5e <- ggplot(moss_data, aes(x = factor(Species), y = Shannon, fill =   
                                factor(Compartment))) +
  geom_boxplot(alpha = 0.9) +
  scale_fill_manual(values = c("salmon4", "darkolivegreen"), labels=c("Brown","Green")) +
 geom_point(position=position_jitterdodge(dodge.width=0.7,seed=0.5),alpha=0.4,show.legend = FALSE) +
  ylab("Bacterial diversity (Shannon index)")+
  facet_wrap(~ Species, scale="free") +
  theme(axis.text.x = element_blank(),
  text = element_text(size = 12),
  axis.title = element_text(size = 13, face = "bold"),
  axis.text = element_text(size = 10),
  legend.position = "bottom",
  legend.title = element_blank(),
  legend.text = element_text(size = 10))
 
 plot5e

#3. Change the name of the legend:
#+ labs(fill = "Compartment")
 plot5f <- ggplot(moss_data, aes(x = factor(Species), y = Shannon, fill =   
                                factor(Compartment))) +
  geom_boxplot(alpha = 0.9) +
  scale_fill_manual(values = c("salmon4", "darkolivegreen"), labels=c("Brown","Green")) +
 geom_point(position=position_jitterdodge(dodge.width=0.7,seed=0.5),alpha=0.4,show.legend = FALSE) +
  ylab("Bacterial diversity (Shannon index)")+
  facet_wrap(~ Species, scale="free") +
  theme(axis.text.x = element_blank(),
  text = element_text(size = 12),
  axis.title = element_text(size = 13, face = "bold"),
  axis.text = element_text(size = 10),
  legend.position = "right",
  legend.text = element_text(size = 10)) +
  labs(fill = "Compartment")
 
 plot5f
 
 #+  labs(x = "Species", fill = "Compartment")
  plot5f <- ggplot(moss_data, aes(x = factor(Species), y = Shannon, fill =   
                                factor(Compartment))) +
  geom_boxplot(alpha = 0.9) +
  scale_fill_manual(values = c("salmon4", "darkolivegreen"), labels=c("Brown","Green")) +
 geom_point(position=position_jitterdodge(dodge.width=0.7,seed=0.5),alpha=0.4,show.legend = FALSE) +
  ylab("Bacterial diversity (Shannon index)")+
  facet_wrap(~ Species, scale="free") +
  theme(axis.text.x = element_blank(),
  text = element_text(size = 12),
  axis.title = element_text(size = 13),
  axis.text = element_text(size = 10),
  legend.position = "right",
  legend.text = element_text(size = 10)) +
   labs(x = "Moss species", fill = "Compartment")
 
 plot5f
 
 #remove gridlines
 
  plot5f <- ggplot(moss_data, aes(x = factor(Species), y = Shannon, fill =   
                                factor(Compartment))) +
  geom_boxplot(alpha = 0.9) +
  scale_fill_manual(values = c("salmon4", "darkolivegreen"), labels=c("Brown","Green")) +
 geom_point(position=position_jitterdodge(dodge.width=0.7,seed=0.5),
            alpha=0.4,show.legend = FALSE) +
  ylab("Bacterial diversity (Shannon index)")+
  facet_wrap(~ Species, scale="free") +
  theme(axis.text.x = element_blank(),
  text = element_text(size = 12),
  axis.title = element_text(size = 13),
  axis.text = element_text(size = 10),
  legend.position = "right",
  legend.text = element_text(size = 10),
  panel.grid.major = element_blank(), # Remove major gridlines
  panel.grid.minor = element_blank(), # Remove minor gridlines
  panel.background = element_rect(fill = "white", colour = "white")) +
  labs(x = "Moss species", fill = "Compartment")
 
 plot5f
 
```

What if we want to add summary statistics?
We need to calculate it outside plot and then add it to the plot as an extra layer
```{r}
install.packages("ggsignif")
library(ggsignif)

#Let's say we have run our test and know the summery statistics
p_values <- data.frame(
  Species = c('Polytrichum piliferum', 'Polytrichum juniperinum','Polytrichum commune','Dicranum undulatum'),
  Compartment = c('Brown', 'Green'), # Replace with actual compartment names if needed
  p_value = c(0.032, 0.045, 0.001, 0.005), # Replace with actual p-values from your tests
  ypos = c(7.8, 7.9) # Replace with the y-axis positions where you want the p-values to appear
)

plot5h <- plot5f +
  geom_text(
    data = p_values,
    aes(x = Species, y = ypos, label = paste0("p = ", p_value)),
    size = 4,
    vjust = 1,
    hjust = -0.6)

plot5h



#MORE complicated example
library(tidyverse)
library(reshape2)
library(ggpubr)

model <- lm (Shannon ~ Compartment + Species, data = moss_data)
comp_pairwise <- emmeans(model, pairwise ~ Compartment |Species, adjust= "tukey")

summary(comp_pairwise) # you get all the p-values and effect sizes here 
p.val.test<-pwpm(comp_pairwise, means = FALSE, flip = TRUE, reverse = TRUE)

# all this code to make the stars
p.val.test <- p.val.test$`Species = Polytrichum juniperinum` # this only looks at ONE HOST AT A TIME. 

p.val.test<-sub("[<>]", "", p.val.test) # is a character 
p.matx<-matrix(as.numeric((p.val.test)),nrow = length(p.val.test[,1]),
               ncol = length(p.val.test[,1])) #if your factor has 5 levels ncol and nrow=5
rownames(p.matx) <- colnames(p.matx) <-colnames(p.val.test)
p.matx[upper.tri(p.matx, diag=FALSE)] <- NA
stat.test<-subset(melt(p.matx),!is.na(value))
names(stat.test)<-c("group1","group2","p.adj")
stat.test[stat.test$p.adj<=0.001,"p.adj.signif"]<-"***"
stat.test[stat.test$p.adj>0.001 & stat.test$p.adj<=0.01,"p.adj.signif"]<-"**"
stat.test[stat.test$p.adj>0.01 & stat.test$p.adj<=0.05,"p.adj.signif"]<-"*"
stat.test[ stat.test$p.adj>0.05,"p.adj.signif"]<-"ns"
test<-mc_tribble(stat.test) # copy & paste the OUTPUT of this line before the line of ggboxplot!!

# makes a tibble table with the p-values, and where they should go 
comp_pval <- tribble(
  ~Species, ~group1, ~group2, ~p.adj, ~p.adj.signif,
  "P_commune","Green","Brown", 1e004,"***", 
  "P_juniperinum","Green","Brown", 0.0445,"*",
  "P_piliferum", "Green", "Brown", 0.0102, "**"
  )

 
 plot5f + stat_pvalue_manual(comp_pval,
                             y.position = max(div_plots$Shannon)+sd(div_plots$Shannon),
                             x.position = div_plots$Species,
                             color = "black",
                             size = 8,
                             bracket.size = 1)
```
Ok, boxplots are nice, but we also want to show a data distribution. How can we do that?
We can use raincloud plots



```{r}
install.packages("ggdist")
library(ggdist)


plot5g <- ggplot(moss_data, aes(x = factor(Species), y = Shannon, fill = factor(Compartment))) +
  # Half-violin plot (or just the "line" of the density)
  stat_halfeye(
    position = position_dodge(width = 0.8),
    adjust = 1.5,  # Adjust this for smoothing
    .width = 0.5  # Control the width of the density plot
  ) +
  # Point data for individual observations
  geom_point(
    position = position_jitterdodge(dodge.width = 0.8),
    alpha = 0.4,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = c("salmon4", "darkolivegreen"), labels = c("Brown", "Green")) +
  ylab("Bacterial diversity (Shannon index)") +
  facet_wrap(~ Species, scale = "free") +
  theme(
    axis.text.x = element_blank(),
    text = element_text(size = 12),
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 10),
    legend.position = "right",
    legend.text = element_text(size = 10),
    panel.grid.major = element_blank(), # Remove major gridlines
    panel.grid.minor = element_blank(), # Remove minor gridlines
    panel.background = element_rect(fill = "white", colour = "white")
  ) +
  labs(x = "Moss species", fill = "Compartment")

plot5g


plot5r <- plot5g +
  geom_text(
    data = p_values,
    aes(x = Species, y = ypos, label = paste0("p = ", p_value)),
    size = 4,
    vjust = 1,
    hjust = -0.6)

plot5r


#what if we had more normally distributed data and more data points
moss_datanew <- read.csv("C:/Users/maria/Dropbox/Postdoc_Sherbrooke/Courses/R course visualization with ggplot/Data/moss_datanew.csv", sep = ",")

plotnew <- ggplot(moss_datanew, aes(x = factor(Species), y = Shannon, fill = factor(Compartment))) +
  # Half-violin plot (or just the "line" of the density)
  stat_halfeye(
    position = position_dodge(width = 0.6),
    adjust = 1.1,  # Adjust this for smoothing
    .width = 0.25  # Control the width of the density plot
  ) +
  geom_boxplot(
    width = 0.2,  # Narrower boxplot width to fit inside the violins
    position = position_dodge(width = 0.6),
    alpha = 0.5
  ) +
  # Point data for individual observations
  geom_point(
    position = position_jitterdodge(dodge.width = 0.6),
    alpha = 0.4,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = c("salmon4", "darkolivegreen"), labels = c("Brown", "Green")) +
  ylab("Bacterial diversity (Shannon index)") +
  facet_wrap(~ Species, scale = "free") +
  theme(
    axis.text.x = element_blank(),
    text = element_text(size = 12),
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 10),
    legend.position = "right",
    legend.text = element_text(size = 10),
    panel.grid.major = element_blank(), # Remove major gridlines
    panel.grid.minor = element_blank(), # Remove minor gridlines
    panel.background = element_rect(fill = "white", colour = "white")
  ) +
  labs(x = "Moss species", fill = "Compartment")

plotnew



```
Exercise
Using this mock dataset of bioassay results create box plots. 

1. Start with a simple boxplot, 
2. Then add colors/fills for Treatments, 
3. Try faceting by Time, 
4. Edit theme

```{r}
bioassay <- read.csv("C:/Users/maria/Dropbox/Postdoc_Sherbrooke/Courses/R course visualization with ggplot/Data/bioassay_data.csv", sep = ",")


# 1. Start with a simple boxplot
simple_boxplot <- ggplot(bioassay, aes(x = Group, y = Response)) +
  geom_boxplot()
simple_boxplot

# 2. Then add colors/fills for Treatments
fill <- ggplot(bioassay, aes(x = Group, y = Response, fill = Group)) +
  geom_boxplot()
fill

# 3. Try faceting by Time
faceted_boxplot <- ggplot(bioassay, aes(x = Group, y = Response, fill = Group)) +
  geom_boxplot() +
  facet_wrap(~ Timepoint)
faceted_boxplot


# 4. Edit theme
# Define custom colors for the treatment groups
install.packages("tidyquant")
library(tidyquant)

beautiful_boxplot <- faceted_boxplot +
    scale_fill_tq() +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold", color = "black"),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.box.background = element_blank(),
    legend.box.margin = margin(0, 0, 0, 0),
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, color = "grey50"),
    axis.text.y = element_text(color = "grey50"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "grey50", fill = NA, size = 1), # Adds panel outline
    panel.spacing = unit(2, "lines") # Adjust the spacing between facets
  ) +
  labs(
    x = "Treatment",
    y = "Value"
  ) +
  guides(fill = guide_legend(title_position = "top", title_hjust = 0.5))

# Display the themed boxplot with panel outlines
print(beautiful_boxplot)
```


